---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# navigm

The `navigm` is an R package that implements a Gaussian graphical spike-and-slab model for estimating associations between variables (nodes). In addition to the primary measurements of node variables, this package allows for incorporating node-level auxiliary variables. The method is tailored to enhance the detection of conditional dependence structure by leveraging and selecting useful node-level information. The package employs a scalable variational Bayes expectation conditional maximisation algorithm to obtain posterior mode of the precision matrix that encodes conditional dependence between nodes and approximated posterior distributions of the remaining model parameters.

The package offers several modeling options, including the choice of incorporating node-level variables and the option to select influential variables. Furthermore, it provides an alternative deterministic inference algorithm, full expectation conditional maximisation, which has been widely used in previous studies.


Reference: Xiaoyue Xi, Helene Ruffieux, 2023. A hierarchical framework for inferring and leveraging node-level information in Bayesian networks. 
  

## Installation
The package can be installed in R by
```{r install, eval=F}
if(!require(remotes)) install.packages("remotes")
remotes::install_github("XiaoyueXI/navigm")
```

For further instructions, please check the (tutorial)[link].



