---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# navigm

The `navigm` is an R package that implements the graphical spike-and-slab in Gaussian graphical models. Besides the primary measurements of variables, this package allows for incorporating node-level auxiliary variables. The method is tailored to improve the detection of conditional independence by leveraging node-level information while selecting useful node-level variables that influence the graph structure. The package implements a scalable variational Bayes expectation-maximization algorithm to obtain the approximated posterior distribution of the model parameters and the posterior mode of precision matrices.

The package offers several modeling options, including whether to incorporate node-level variables and whether to make the selection of variables. Additionally, it provides deterministic inference options, such as full expectation maximization and variational Bayes expectation-maximization algorithms.

Reference: Xiaoyue Xi, Sylvia Richardson, Helene Ruffieux, 2023. A hierarchical framework for inferring and leveraging node-level information in Bayesian networks. 
  

## Installation
The package can be installed in R by
```{r install, eval=F}
if(!require(remotes)) install.packages("remotes")
remotes::install_github("XiaoyueXI/navigm")
```


```{r personal install, eval=T, include=F}
devtools::load_all('~/navigm/')
```

