---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# navigm

`navigm` is an R package that implements a Gaussian graphical spike-and-slab model for estimating 
associations between variables (nodes). Alongside the primary node measurements,
this package permits the inclusion of supplementary node-level auxiliary variables. 
This method is tailored to enhance the identification of conditional dependence
structure by leveraging these variables and selecting useful information from them. 
Employing a scalable variational Bayes expectation conditional maximisation algorithm,
the package attains the posterior mode of the precision matrix that pinpoints the conditional dependence between nodes,
along with approximated posterior distributions of the remaining model parameters.

The package offers a variety of modelling options,
including the decision to integrate node-level variables as well as the flexibility to select influential variables. 
Furthermore, it provides an alternative deterministic
inference algorithm, full expectation conditional maximisation, which
has been widely used in previous studies.

Reference: Xiaoyue Xi, Helene Ruffieux, 2023. A hierarchical framework
for inferring and leveraging node-level information in Bayesian
networks.
  

## Installation
The package can be installed in R by
```{r install, eval=F}
if(!require(remotes)) install.packages("remotes")
remotes::install_github("XiaoyueXI/navigm")
```

For further instructions, please check the (tutorial)[link].



